---
title: "Real-time Model of Probability Matrix of Soccer Games"
author: "Michael Yan"
date: "June 28, 2016"
output:
  rmdformats::readthedown:
    highlight: kate
---


```{r knitr_init, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(max.print="75")
opts_chunk$set(#echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=80)
```


## Preliminary Processing

Load packages and data that going to be used. I aslo define a theme function for customized ggplots.

```{r, message=F, warning=F}
.packages = c("dplyr", "caret", "scales", "gridExtra","rmdformats","knitr", "corrplot")
.inst <- .packages %in% installed.packages()
if(length(.packages[!.inst]) > 0) install.packages(.packages[!.inst],repos = 'http://cran.us.r-project.org')
lapply(.packages, require, character.only=TRUE)

rawdata <- read.csv("TrainingData_Sample2.csv", stringsAsFactors=FALSE)
mycolor <- c(rep(c("#FF5A5F","#FFB400", "#007A87", "#8CE071", "#7B0051", "#00D1C1",
                    "#FFAA91", "#B4A76C", "#9CA299", "#565A5C", "#00A04B","#E54C20"), 100))
theme_custom <- function(base_size = 11, base_family = "Franklin Gothic Medium") {
  theme_grey(base_size = base_size, base_family = base_family) %+replace%
    theme(
      line =              element_line(colour = '#DADADA', size = 0.75, 
                                       linetype = 1, lineend = "butt"),
      rect =              element_rect(fill = "#F0F0F0", colour = "#F0F0F0", 
                                       size = 0.5, linetype = 1),
      text =              element_text(family = base_family, face = "plain",
                                       colour = "#656565", size = base_size,
                                       hjust = 0.5, vjust = 0.5, angle = 0, 
                                       lineheight = 0.9,margin = margin(), debug = FALSE),
      plot.title =        element_text(size = rel(1.5), family = base_family , 
                                       face = 'bold', hjust = -0.05, 
                                       vjust = 1.5, colour = '#3B3B3B'),
      axis.title.x =      element_blank(),
      axis.title.y =      element_blank(),
      axis.text =         element_text(),
      axis.ticks =        element_line(),
      panel.grid.major =  element_line(),
      panel.grid.minor =  element_blank(),
      plot.background =   element_rect(),
      panel.background =  element_rect(),
      legend.key =        element_rect(colour = '#DADADA'),
      legend.position = 'none',
      complete = TRUE
    )
}
```

By looking at the discriptive statistics of the dataset, we find that *NeutralField* is constant to 0. We can remove this variable from the dataset. Current score, current red cards nad remaining goals of both teams are calculated from the raw dataset.

```{r, message=F, warning=F}
summary(rawdata$NeutralField)
cleandata <- rawdata %>%
  select(-NeutralField) %>%
  mutate(RC_H = (RCTotal + RCDiff)/2, RC_A = (RCTotal - RCDiff)/2, 
         CScore_H = (TotalScore + ScoreDiff)/2, CScore_A = (TotalScore - ScoreDiff)/2,
         RMScore_H = rbS.FinalScoreH - CScore_H, RMScore_A = rbS.FinalScoreA - CScore_A)
```

It is widely accepted in the literatures that Poisson distribution should capture the distribution of home and away goals. Lets take a look at the histogram of Home and Aways goals of our dataset alongside with the simulated goals from Poisson distribution using the mean of goals from our dataset. Both home and away chart are relatively close to the simulated one, our notion is supported.

```{r, echo = F, message=F, warning=F, fig.height= 6, fig.width= 10}
p1 = ggplot(cleandata %>% group_by(ID) %>% summarise(home = max(rbS.FinalScoreH), away = max(rbS.FinalScoreA)), aes(home)) + 
  geom_histogram(aes(y = ..density..),binwidth = 1,fill = mycolor[1], color = mycolor[7]) +
  ggtitle("Histogram of number of Home Goals") + 
  theme_custom()

p2 = ggplot(cleandata %>% group_by(ID) %>% summarise(home = max(rbS.FinalScoreH), away = max(rbS.FinalScoreA)), aes(away)) + 
  geom_histogram(aes(y = ..density..),binwidth = 1,fill = mycolor[1], color = mycolor[7]) +
  ggtitle("Histogram of number of Away Goals") + 
  theme_custom()

a1 <- cleandata %>% group_by(ID) %>% summarise(home = max(rbS.FinalScoreH))
b1 <- as.data.frame(rpois(99999,mean(a1$home)))
names(b1)[1] = "simu"

a2 = cleandata %>% group_by(ID) %>% summarise(away = max(rbS.FinalScoreA))
b2 = as.data.frame(rpois(99999,mean(a2$away)))
names(b2)[1] = "simu"

p3 = ggplot(b1, aes(simu)) + 
  geom_histogram(aes(y = ..density..),binwidth = 1,fill = mycolor[1], color = mycolor[7]) +
  ggtitle("Histogram of Simulated Home Goals") + 
  theme_custom()

p4 = ggplot(b2, aes(simu)) + 
  geom_histogram(aes(y = ..density..),binwidth = 1,fill = mycolor[1], color = mycolor[7]) +
  ggtitle("Histogram of Simulated Away Goals") + 
  theme_custom()

grid.arrange(p1,p2,p3,p4, nrow =2)
```

However, in this question we are not trying to predict the game result, but to perdict the result of the remainder of the game. *RMScore_H* and *RMScore_A* are calculated using *rbS.FinalScore* subtracted by current score. Generating the same histogram for *RMScore_H* and *RMScore_A* as above, we can assume that goals in the remainder of the game is also following Poisson distribution.

```{r, echo = F,message=F, warning=F, fig.height= 6, fig.width= 10}
p1 = ggplot(cleandata, aes(RMScore_H)) + 
  geom_histogram(aes(y = ..density..),binwidth = 1,fill = mycolor[1], color = mycolor[7]) +
  ggtitle("Histogram of Remaining Home Goals") + 
  theme_custom()

p2 = ggplot(cleandata, aes(RMScore_A)) + 
  geom_histogram(aes(y = ..density..),binwidth = 1,fill = mycolor[1], color = mycolor[7]) +
  ggtitle("Histogram of Remaining Away Goals") + 
  theme_custom()

b1 = as.data.frame(rpois(99999,mean(cleandata$RMScore_H)))
names(b1)[1] = "simu"

b2 = as.data.frame(rpois(99999,mean(cleandata$RMScore_A)))
names(b2)[1] = "simu"

p3 = ggplot(b1, aes(simu)) + 
  geom_histogram(aes(y = ..density..),binwidth = 1,fill = mycolor[1], color = mycolor[7]) +
  ggtitle("Simulated Remaining Home Goals") + 
  theme_custom()

p4 = ggplot(b2, aes(simu)) + 
  geom_histogram(aes(y = ..density..),binwidth = 1,fill = mycolor[1], color = mycolor[7]) +
  ggtitle("Simulated Remaining Away Goals") + 
  theme_custom()

grid.arrange(p1,p2,p3,p4, nrow =2)



```


Let's take a look at the distribution of the time of goals

```{r, message=F, warning=F}
goaltime <- cleandata %>% 
  group_by(ID) %>%
  mutate(gtime = TotalScore - lag(TotalScore)) %>%
  select(ID, Minute, TotalScore, gtime, HasPenaltyShootout) %>%
  filter(gtime == 1)

ggplot(goaltime, aes(Minute)) + 
  geom_histogram(aes(y = ..density..),binwidth = 1,fill = mycolor[1], color = mycolor[2]) +
  geom_density(color=mycolor[3],size = 1.5) +
    ggtitle("Histogram of Goals") + 
  theme_custom()
```
We can see that the distribution of the goals with 91 minutes of the match tends to gradually increase. This might due to the player tiredness, or during the early staged of the game the players are not fully warmed up yet.

## Model Methodology

After we made the assumption that the number of goals in a game is following poisson distribution, The goal outcome of a remainder of the match between home team away team in match $k$ is modeled as:
$$HomeGoals_k \sim Poisson(\lambda_{k,t})$$
$$AwayGoals_k \sim Poisson(\mu_{k,t})$$
The parameters of the Poisson distribution are define as 
$$log(\lambda_{k,t}) = \beta_0 + \beta_1*factor_1 + \beta_2*factor_2 + \cdots +  \lambda_{xy}$$
$$log(\mu_{k,t}) = \gamma_0 + \gamma_1*factor_1 + \gamma_2*factor_2 + \cdots + \mu_{xy}$$
where $\beta_{xy}$ is a piecewise function:
$$\lambda_{xy} = \begin{cases} 
      0 & x = y \\
      \lambda_{10} & x > y \\
      \lambda_{01} & x < y 
   \end{cases}
$$
$x$ and $y$ are home goals and away goals respectively.$\mu_{xy}$ is defined similarly. This parameter is based on the assumption that current score will affect game intensity. For example, teams may try to defend a lead or to restore equality if the are ahead or behind respectively.
The our problem became a poisson regression problem with the link function defined as $log(\lambda_{k,t})$. The likelihood contribution from each minutes within a match are summarized and since scores between matchs are assumed to be independent, so the overall log likelihood will be calculated by taking the summation over matches. After derived log-likelihood function, we can use R's *nlm()* or *optim()* functions to get parameter estimations.

## Variable Selection

We have over 30 variables in the dataset and intuitively we shouldn't put them all into the regression modelbecause this might overfit the model. Also, many of the independent variable clearly have certain degree of correlation. For example, Danger Zone Attack is definetly associated with Corner Kicks. Therefore, some preliminary work need to be done in order to reduce the variable list. Let's take a look at the pearson correlation matrix of the dataset:

```{r, message=F, warning=F,fig.height= 10, fig.width= 10}
corr = round(cor(select(cleandata, -League, -DrawProb, -HasPenaltyShootout)),3)
corrplot(corr)
```

Clearly we saw high correlations between in-game stats such as SoG, DAT, CR,etc. It is normal to pick variables that has higher correlation with our dependent variable *RMScore_H* and *RMScore_A* and has lower cross-correlation between each other. Based on this cratiea, *MeanH, MeanA, Minute, DAT, SoG, YC, FK* are selected variable going into log regression model





##

